Water discharge quantity has a variable impact on sediment transport capacity in subglacial channels

### 2023-10-02 Delaneyetal2023b.tex

A few basic hydraulic relations are used in the following.  The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi$ the head gradient and friction factor $f$.  The shear stress and stream power are, respectively,
\begin{equation}
    \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%
Sediment discharge is given by the MPM, EH and Bagnold formulations as, respectively,
\begin{equation}
  Q_s \propto w\, \tau^{3/2}, \quad Q_s \propto w\, \tau^{5/2}, \quad Q_s \propto w\, \Omega^{3/2}
\end{equation}
for conditions well above the transport onset threshold.

The river is assumed to have a width $W$ much greater than its depth $H$, such that $D_h\approx 4H$, and to have a constant head gradient ($\Psi$) given by the topography.  Further, we assume that its width can be approximated by a relation $w \propto Q^\alpha$ with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$, $\alpha=0$ or $\alpha=1$ correspond a river of constant width or depth, respectively.
%
For a steady state R-channel, we assume that it has a constant $\Psi$ (approximated by the gradient of the Shreve \citeyear{shreve1972} potential) and that $S$ adjusts such that it is in steady state corresponding to $\Psi$ and $Q$.  %Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
%
For pipe flow, which occurs when a R-channel is subjected to rapid discharge variations such that it cannot adjust its size significantly,  we assume that the cross-sectional area $S$ is fixed and that $\Psi$ adjusts to meet the specified $Q$.  For both R-channel and pipe flow, we assume that $D_h \propto S^{1/2}$.

@@

\mauro{Bagnold gives pretty crazy results. Check again!}

Table~\ref{tab:Qs2} shows that the proxy $W\tau$ as well as the MPM and EH sediment transport formulas give roughly a similar scaling of $Q_s$ on $\Psi$ or $S$ and on $Q$.  The total transport formula EH giving a stronger dependence on all variables as should be expected duto to the larger exponent on $\tau$ of 2.5 versus 1.5 for the others.  The exponent for the width scaling $\alpha$ only impacts the EH relation, with its value almost certainly around or bigger than 0.5 for a braided proglacial stream.
%
The sediment discharge in the steady state R-channel scales very similarly to the river case, and almost identically for the $\alpha=0.5$ case.  However, note that the head gradients $\Psi$ are likely much higher for comparable $Q$ in a glacial setting than a river setting, as is well described in \citeA{alley1997}, and thus transport rates may still be much higher in a steady state R-channel.
%
Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.  In such cases they operate more like a pipe of fixed cross-section.  Table~\ref{tab:Qs2} shows that then the sediment transport scales much more severely with discharge, with the exponent of $Q$ being between 2 and 5, than in the other two regimes when that exponent is at most 1.3.  Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause very high sediment transport conditions.

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes: river, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
  \small
\begin{tabular}{llllll}
Regime & Fixed & Determined via D-W & Additional relations & Shear stress & Stream p.\\
 &  &  &  & \(\tau \propto\) & \(\Omega \propto\)\\
\hline
River & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \(W\,\propto Q^\alpha\), \(S=WH\), \(D_h=H\) & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q \Psi\)\\
R-ch. & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{Sediment transport proxy ($W\tau$) and rates for the three considered different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{Bagnold1980}.
    }
  \label{tab:Qs}
\begin{tabular}{lllll}
 & Width*\(\tau\) & MPM & EH & Bagnold\\
 & \(W\, \tau\) & \(Q_s \propto W\, \tau^{3/2}\) & \(Q_s \propto W\, \tau^{5/2}\) & \(Q_s \propto W\, \Omega^{3/2} H^{-2/3}\)\\
\hline
River & \(f^{1/3}\, Q^{2/3+\alpha/3}\,  \Psi^{2/3}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{5/6}\, Q^{5/3 - 2\alpha/3} \, \Psi^{5/3}\) & \(f^{-2/9}\, Q^{13\alpha/9 - 19/18} \, \Psi^{31/18}\)\\
R-ch. & \(f^{2/5}\, Q^{4/5} \, \Psi^{3/5}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{7/10}\, Q^{7/5}\, \Psi^{9/5}\) & \(f^{1/15}\, Q^{49/30}\, \Psi^{43/30}\)\\
Pipe & \(f \, Q^2 \, S^{-1}\) & \(f^{3/2}\, Q^3 \, S^{-5/2}\) & \(f^{5/2}\, Q^5\, S^{-9/2}\) & \(f^{3/2} \, Q^{9/2} \, S^{-43/12}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{Sediment transport proxy ($W\tau$) and rates for the three considered different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{Bagnold1980}.
    }
  \label{tab:Qs2}
\begin{tabular}{lllll}
 & Width*\(\tau\) & MPM & EH & Bagnold\\
 & \(W\, \tau\) & \(Q_s \propto W\, \tau^{3/2}\) & \(Q_s \propto W\, \tau^{5/2}\) & \(Q_s \propto W\, \Omega^{3/2} H^{-2/3}\)\\
\hline
River (\(\alpha=0\)) & \(f^{0.7}\, Q^{0.7}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{-1.1} \, \Psi^{1.7}\)\\
River (\(\alpha=0.5\)) & \(f^{0.7}\, Q^{0.8}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\,\Psi\) & \(f^{0.8}\, Q^{1.3} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{-0.3} \, \Psi^{1.7}\)\\
River (\(\alpha=1\)) & \(f^{0.7}\, Q^{1.0}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{0.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{0.4} \, \Psi^{1.7}\)\\[3pt]
R-ch. & \(f^{0.4}\, Q^{0.8} \, \Psi^{0.6}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.7}\, Q^{1.4}\, \Psi^{1.8}\) & \(f^{0.07}\, Q^{1.6}\, \Psi^{1.4}\)\\
Pipe & \(f \quad\, Q^{2\phantom{.0}} \, S^{-1}\) & \(f^{1.5}\, Q^3 \, S^{-2.5}\) & \(f^{2.5}\, Q^{5\phantom{.0}}\, S^{-4.5}\) & \(f^{1.5} \, Q^{4.5} \, S^{-3.6}\)\\
\end{tabular}
\end{table}

>>>>

A few basic hydraulic relations are used in the following.  The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi$ the head gradient and friction factor $f$.  The shear stress and stream power are, respectively,
\begin{equation}
    \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%
Sediment discharge is given by the MPM, EH and Bagnold formulations as, respectively,
\begin{equation}
  Q_s \propto W\, \tau^{3/2}, \quad Q_s \propto W\, \tau^{5/2}, \quad Q_s \propto W \left(\frac{\Omega}{W}\right)^{3/2} H^{-2/3}
\end{equation}
for conditions well above the transport onset threshold.

The river is assumed to have a width $W$ much greater than its depth $H$, such that $D_h\approx 4H$, and to have a constant head gradient ($\Psi$) given by the topography.  Further, we assume that its width can be approximated by a relation $W \propto Q^\alpha$ with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel), $\alpha=0$ or $\alpha=1$ correspond a river of constant width (a slot canyon) or depth (no natural equivalent), respectively, i.e. the two end members.
%
For a steady state R-channel, we assume that it has a constant $\Psi$ (approximated by the gradient of the Shreve \citeyear{shreve1972} potential) and that $S$ adjusts such that it is in steady state corresponding to $\Psi$ and $Q$.  %Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
%
For pipe flow, which occurs when a R-channel is subjected to rapid discharge variations such that it cannot adjust its size significantly,  we assume that the cross-sectional area $S$ is fixed and that $\Psi$ adjusts to meet the specified $Q$.  For both R-channel and pipe flow, we assume that $D_h \propto S^{1/2}$.

@@

Table~\ref{tab:Qs2} shows that the proxy $W\tau$ as well as the MPM, EH and Bagnold sediment transport formulas give roughly a similar scaling of $Q_s$ on $\Psi$ or $S$ and on $Q$.  Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime, which seems rather unexpected.  The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to to the larger exponent on $\tau$ of 2.5 versus 1.5 for the others.  Albeit, for pipe-flow Bagnold is close to EH. Conversely, the proxy used within this publication scales only as $Q^2$ for pipe-flow, whereas the others scale at least with $Q^3$.  The exponent for the width scaling $\alpha$ only impacts the EH relation in any meaningful way; but for the value of $\alpha$ around 0.5, as is suitable for a braided proglacial stream and indeed most rivers, the exponent on $Q$ is only slightly above what the other relations give.

The sediment discharge in the steady state R-channel scales very similarly to the river case for all relations and virtually identically for the $\alpha=0.5$ case.  However, note that the head gradients $\Psi$ are likely higher for comparable $Q$ in a ice-sheet marginal or alpine glacier setting than in a river setting, as is well described in \citeA{alley1997}, and thus transport rates may still be much higher in a steady state R-channel.

Furthermore, R-channels will rarely operate in steady state as variations in discharge, in particular on the diurnal timescale or the timescale of severe rain or melting events, are too fast for such a channel to reach a steady state.  In such cases they operate more like a pipe of fixed cross-section.  Table~\ref{tab:Qs2} shows that in such a situation, the sediment transport scales much more severely with discharge, with the exponent of $Q$ being likely between 3 and 5, compared to other two regimes when that exponent is at most 1.3.  Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause very high sediment transport conditions.

\begin{table}
  \caption{Relations for hydraulic variables for the three drainage regimes: river, R-channel and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \label{tab:eqs1}
  \small
\begin{tabular}{llllll}
Regime & Fixed & Determined via D-W & Additional relations & Shear stress & Stream p.\\
 &  &  &  & \(\tau \propto\) & \(\Omega \propto\)\\
\hline
River & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \(W\,\propto Q^\alpha\), \(S=WH\), \(D_h=H\) & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q \Psi\)\\
R-ch. & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto W \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q \Psi\)\\
Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto W \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f Q^3 S^{-5/2}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{Sediment transport proxy ($W\tau$) and rates for the three considered different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{Bagnold1980}.
    }
  \label{tab:Qs}
\begin{tabular}{lllll}
 & Width*\(\tau\) & MPM & EH & Bagnold\\
 & \(W\, \tau\) & \(Q_s \propto W\, \tau^{3/2}\) & \(Q_s \propto W\, \tau^{5/2}\) & \(Q_s \propto W^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
\hline
River & \(f^{1/3}\, Q^{2/3+\alpha/3}\,  \Psi^{2/3}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{5/6}\, Q^{5/3 - 2\alpha/3} \, \Psi^{5/3}\) & \(f^{-2/9}\, Q^{19/18-\alpha/18} \, \Psi^{31/18}\)\\
R-ch. & \(f^{2/5}\, Q^{4/5} \, \Psi^{3/5}\) & \(f^{1/2}\, Q \, \Psi\) & \(f^{7/10}\, Q^{7/5}\, \Psi^{9/5}\) & \(f^{-7/30}\, Q^{31/30}\, \Psi^{26/15}\)\\
Pipe & \(f \, Q^2 \, S^{-1}\) & \(f^{3/2}\, Q^3 \, S^{-5/2}\) & \(f^{5/2}\, Q^5\, S^{-9/2}\) & \(f^{3/2} \, Q^{9/2} \, S^{-14/3}\)\\
\end{tabular}
\end{table}

\begin{table}
  \caption{As Table~\ref{tab:Qs} but with the fractional exponents stated as rounded decimal numbers.  For the River regime the exponents are for displayed for the likely width-exponent $\alpha=1$, as well as its end members 0 (slot canyon) and 1 (only width increases and not depth).
    }
  \label{tab:Qs2}
\begin{tabular}{lllll}
 & Width*\(\tau\) & MPM & EH & Bagnold\\
 & \(W\, \tau\) & \(Q_s \propto W\, \tau^{3/2}\) & \(Q_s \propto W\, \tau^{5/2}\) & \(Q_s \propto W^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
\hline
River (\(\alpha=0\)) & \(f^{0.7}\, Q^{0.7}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.1} \, \Psi^{1.7}\)\\
River (\(\alpha=0.5\)) & \(f^{0.7}\, Q^{0.8}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.3} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
River (\(\alpha=1\)) & \(f^{0.7}\, Q^{1.0}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{0.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\[3pt]
R-ch. & \(f^{0.4}\, Q^{0.8} \, \Psi^{0.6}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.7}\, Q^{1.4}\, \Psi^{1.8}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
Pipe & \(f \,\quad Q^{2\phantom{.0}} \, S^{-1}\) & \(f^{1.5}\, Q^3 \, S^{-2.5}\) & \(f^{2.5}\, Q^{5\phantom{.0}}\, S^{-4.5}\) & \(f^{1.5} \,\,\,\,\, Q^{4.5} \, S^{-4.7}\)\\
\end{tabular}
\end{table}
